---
title: '2020FA MATH 1051H: Lecture #05'
output:
  ioslides_presentation:
    css: ../style.css
    fig_caption: yes
    font-family: Lato Semibold
    font-import: http://fonts.googleapis.com/css?family=Lato
    widescreen: yes
  beamer_presentation: default
always_allow_html: true
---
<style>
citation {
  font-size: 4px;
}
</style>

<!--  Version 1.0-0

      This version of the slides is taken from Mine Çetinkaya-Rundel's lecture slides
      posted on OpenIntro.org in .pptx and .gdslides format, and moved to Rmd. Originally done
      as a translation for github.com/wesleyburr/introstat_random/. Some material added throughout,
      and some things cleaned up.
      
      A large part of the HTML/CSS formatting is janky, and could be cleaned up. Feel free to issue a 
      pull request if you love HTML and CSS and want to fix this up.
      
      - wburr, Sept 11, 2018
-->

# Statistics and Summary Plots

## Today's Review

```{r, echo = FALSE}
set.seed(12)
```

In today's lecture, we are going to review the ideas of Chapters 2.1 and
2.2 of our textbook, with several worked examples. The code demonstrated
today will be gone over again in Workshops #03 and #04, in different
ways, so that hopefully the combination will allow you to grasp how to work
with things.

## To Start ... Data

To work with data in R, it must be organized in some fashion. The two
objects we will use the most to do this are:

* **The Vector**: a comma-separated list of elements
* **The Data Frame**: a bunch of vectors, organized as columns in a spreadsheet-like object

The data.frame is more complicated, and I'll show you a quick demo of it today
as we extract data, but we'll be **working** with the vector for your assignments
for a while.

## Example 1: Simulated Data

Our first example will use simulated data to explore the ideas. We will
generate 100 randomly selected integers between 1 and 1000. This
is done using the **sample()** function.

```{r}
dat1 <- sample(x = seq(from = 1, to = 1000, by = 1),
               size = 100,
               replace = TRUE)
str(dat1)
```

Now that we have the data, let's summarize it.

## Example 1: Summary Statistics

```{r}
summary(dat1)
sd(dat1)
```

From this summary we see that the sample of 100 observations ranges
from 13 to 999, with median 451.5 (there are an even number of observations)
and mean 473.2. The mean is larger than the median: what does this imply
might be happening?

## Example 1: Summary Plots

```{r}
par(mfrow = c(1, 2))
hist(dat1, breaks = 20)
boxplot(dat1, xlab = "Simulated Data")
```

## Example 1: What Can We Say?

From the two plots and the summary statistics, we might say:

* the data is either uniform or unimodal
* there is a very slight rightward skew, but not a strong one
* there are no obvious outliers
* 50% of the data is between 174 and 754.2

## Example 2: Global Population

The World Health Organization Global Tuberculosis Report contains information
on the cases of tuberculosis worldwide. It also include countries' names and
their populations, which is what we will look at.

```{r}
library("tidyr")
data(who)
data(population)
dat2 <- population$population
str(dat2)
```

## Example 2: Global Population

The population data.frame consists of 4,060 observations of country+year+population. 
We extracted only the population numbers, so we now have 4,060 populations for
these countries over time.

## Example 2: Summary Statistics

```{r}
summary(dat2)
sd(dat2)
```
In this summary, things are presented in scientific notation: any big number (or small
number) in R will be presented this way. We see that the 4,060 observations range
from 1.129e+03 (1,129) to 1.386e+09 (1,386,000,000). Do you know which countries
these might be? 

The mean is clearly **much** greater than the median here (almost 10x larger).

## Example 2: Summary Plots

```{r}
hist(dat2, breaks = 40)
```

## Example 2: Summary Plots

```{r, warning = FALSE}
boxplot(dat2, xlab = "Global Population Data")
```

## Example 2: What Can We Say?

From the two plots and the summary statistics, we might say:

* the data has a strong right skew
* there are obvious outliers
* 50% of the data is between 602,900 and 18,550,000
* 25% of the data is greater than 18,550,000 

## Example 2: What's Happening?

There's too big a range in the data! 25% of the data is above 18 million,
meaning 75% of it is below 18 million ... but above 18 million are many of the countries
you know of (e.g., Canada, USA, Australia, Brazil, Russia, China, India, Germany, France,
the UK, etc.). 

So this data is very odd ... let's take a closer look.

## Example 2: Advanced (Bottom 75%)

I don't expect you to follow the code, so I've hidden it here. What I'm going to 
do is look at only the observations that are less than 19 million (the bottom 75%)
alone, and then the upper 25%.

```{r, echo = FALSE}
dat2b <- dat2[dat2 < 19000000]
dat2c <- dat2[dat2 >= 19000000]
hist(dat2b, main = "Bottom 75% of Country Data", breaks = 19*2)
```

## Example 2: Advanced (Bottom 75%)

```{r, echo = FALSE, warning = FALSE}
boxplot(dat2b, xlab = "Bottom 75% of Country Data")
```

## Example 2: Advanced (Top 25%)

So what does this show? We **still** have a huge number of really small observations,
easily 1/4 of the entire data set, and then a very long right tail. What about the upper 25%?

```{r, echo = FALSE, warning = FALSE}
hist(dat2c, main = "Upper 25% of Country Data", breaks = 2750)
```

## Example 2: Advanced (Top 25%)

So the histogram shows that after the bottom 75%, things start to be very few cases
in each category - the bins of the two plots are the same width, but things are far
more spaced out on the upper end.

```{r, echo = FALSE, warning = FALSE}
boxplot(dat2c, xlab = "Upper 25% of Country Data")
```

## Example 2: What Can We Say?

From our expanded analysis:

* the data has a strong right skew, with lots of outliers
* 25% of the data is greater than 18,550,000, but is very spread out
* even the 75% of the data that is less than 18,550,000 is very strongly
right-skewed

## Example 3: Traffic Fatalities

There is an Applied Econometrics package in R which contains some interesting
data. One of the sets is traffic fatalities for the "lower 48" United States
states for the 7 years from 1982 to 1988 ($48 \times 7 = 336$).

```{r, message = FALSE, warning = FALSE}
library("AER")
data("Fatalities")
dat3 <- Fatalities$fatal
str(dat3)
```

## Example 3: Summary Statistics

```{r}
summary(dat3)
sd(dat3)
```

In this summary, the mean is clearly much greater than the median, so
we expect a right skew. The range is quite large as well: a minimum of 79
fatalities in a given year and state, and a maximum of 5504 fatalities
in another state and a (not necessarily the same) year. 

## Example 3: Summary Plots

```{r}
hist(dat3, breaks = 20)
```

## Example 3: Summary Plots

```{r, warning = FALSE}
boxplot(dat3, xlab = "Total Traffic Fatalities")
```

## Example 3: What Can We Say?

* left truncated - definitely not symmetric
* clear right skew
* many outliers (above about 2500)
* unimodal

## Example 3: Comparing IQR and SD

```{r}
sd(dat3)
IQR(dat3)
```

Notice how the IQR is quite a bit smaller than the standard deviation?
This is the **robustness** discussion from Lecture 04.

# Summary

## What is the point?

When you get a data set, and you've never seen it before, you need to explore it.
The demonstrations today (and on Tuesday and Monday) should give you a hint as to
how to start doing that. There are many possible visualizations, especially when
there are multiple variables, but for single variables, the key things to check 
are:

* summary statistics: mean, median, quantiles, IQR, sd
* the histogram (possibly experiment with multiple breakpoints)
* the boxplot (to see outliers and skew)

## Next?

In Workshops #03 and #04, we'll start unpacking these skills, and giving you the
chance to learn them well, and practice practice practice. You should expect to use
these skills on R Assignments #2, 3 and 4.